---
title: "Assignment 1"
author: "Vinh Nguyen (ID 1029531)"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---


1. Five fair coins are tossed (i.e., the probability of tail and head equals 0.5 for each coin)
and the number of tails, T is counted. Find the conditional probability that T ≥ 1
given that at least one coin shows head.  
H: number of heads  
H + T = 5 (5 tosses)  
P(at least one coin shows head) = P(H >= 1) = P(T <= 4)  
**P(T ≥ 1|T <= 4)** = P(T ≥ 1 and T <= 4) / P(T <= 4) = P(1 <= T <= 4) / P(T <= 4)  
T ~ Bi(n=5, p=0.5)
```{r}
p1 <- sum(dbinom(1:4, 5, 0.5)) # P(1 <= T <= 4)
p2 <- pbinom(4, size=5, prob=0.5) # P(T <= 4)
"P(T ≥ 1|T <= 4)"
p1/p2
```

2. Birthday paradox  
(a) Consider a group of 3 students. Each student has a birthday that can be any one
of the days numbered 1, 2, 3, ..., 365. What is the probability that none of them
have the same birthday with each other?  
Student 1: can take 365 possible birthdays  
Student 2: can take 364 possible birthdays  
Student 3: can take 363 possible birthdays  
All possibilites: 365 x 365 x 365  
P(no same birthday) = (365 x 364 x 363) / (365 x 365 x 365)  
```{r}
"P(no same birthday)"
(365*364*363) / 365^3
```

(b) Consider a group of 23 students. Each student has a birthday that can be any
one of the days numbered 1, 2, 3, ..., 365.  
i. What is the probability that none of them have the same birthday with each
other?  
Student 1: can take 365 possible birthdays  
Student 2: can take (365-1) possible birthdays  
...  
Student 23: can take (365-22) possible birthdays  
All possibilites: 365^23  
```{r}
"P(no same birthday)"
p_null <- prod((365-22):365) / 365^23
p_null
```

ii. What is the probability that some of them have the same birthday with each
other? Is this probability greater than 0.5?  
P(some same birthdays) = 1 - P(no same birthday)  
```{r}
"P(some same birthdays)"
p_some <- 1 - p_null
p_some
"Is this probability greater than 0.5?"
p_some > 0.5
```











