---
title: "Assignment 1"
author: "Vinh Nguyen (ID 1029531)"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---


1. Five fair coins are tossed (i.e., the probability of tail and head equals 0.5 for each coin)
and the number of tails, T is counted. Find the conditional probability that T ≥ 1
given that at least one coin shows head.  
H: number of heads  
H + T = 5 (5 tosses)  
P(at least one coin shows head) = P(H >= 1) = P(T <= 4)  
**P(T ≥ 1|T <= 4)** = P(T ≥ 1 and T <= 4) / P(T <= 4) = P(1 <= T <= 4) / P(T <= 4)  
T ~ Bi(n=5, p=0.5)
```{r}
p1 <- sum(dbinom(1:4, 5, 0.5)) # P(1 <= T <= 4)
p2 <- pbinom(4, size=5, prob=0.5) # P(T <= 4)
"P(T ≥ 1|T <= 4)"
p1/p2
```

2. Birthday paradox  
(a) Consider a group of 3 students. Each student has a birthday that can be any one
of the days numbered 1, 2, 3, ..., 365. What is the probability that none of them
have the same birthday with each other?  
Student 1: can take 365 possible birthdays  
Student 2: can take 364 possible birthdays  
Student 3: can take 363 possible birthdays  
All possibilites: 365 x 365 x 365  
P(no same birthday) = (365 x 364 x 363) / (365 x 365 x 365)  
```{r}
"P(no same birthday)"
(365*364*363) / 365^3
```

(b) Consider a group of 23 students. Each student has a birthday that can be any
one of the days numbered 1, 2, 3, ..., 365.  
i. What is the probability that none of them have the same birthday with each
other?  
Student 1: can take 365 possible birthdays  
Student 2: can take (365-1) possible birthdays  
...  
Student 23: can take (365-22) possible birthdays  
All possibilites: 365^23  
**P(no same birthday)** = (365 x 364 x ... x 343) / 365^23  
```{r}
"P(no same birthday)"
p_null <- prod((365-22):365) / 365^23
p_null
```

ii. What is the probability that some of them have the same birthday with each
other? Is this probability greater than 0.5?  
P(some same birthdays) = 1 - P(no same birthday)  
```{r}
"P(some same birthdays)"
p_some <- 1 - p_null
p_some
"Is this probability greater than 0.5?"
p_some > 0.5
```

3. It is known from experience that in a certain industry 60 percent of all labor-management
disputes are over wages, 15 percent are over working conditions, and 25 percent are
over fringe issues. Also, 45 percent of the disputes over wages are resolved without
strikes, 70 percent of the disputes over working conditions are resolved without strikes,
and 40 percent of the disputes over fringe issues are resolved without strikes. What
is the probability that a labour-management dispute in this industry will be resolved
without a strike?  
Let A1 = event where a dispute is over wages  
A2 = event where a dispute is over working conditions  
A3 = event where a dispute is over fringe issues  
B = event where a dispute is resolved without a strike  
Applying the Law of Total Probability:  
P(B) = P(A1)P(B|A1) + P(A2)P(B|A2) + P(A3)P(B|A3)  
```{r}
"Probability that a labour-management dispute in this industry will be resolved without a strike"
0.6*0.45 + 0.15*0.7 + 0.25*0.4
```

4. In a certain community, 8 percent of all adults have Type 2 diabetes. If a health service
in this community correctly diagnoses 95 percent of adults with diabetes as having
the disease and incorrectly diagnoses 2 percent adults without diabetes as having the
disease, find the probabilities that  
(a) the community health service will diagnose an adult as having diabetes;  
Let X = 1: event that an adult has Type 2 diabetes (P(X=1) = 0.08)  
X = 0: event that an adult does not have Type 2 diabetes (P(X=0) = 0.92)  
Y = 1: event that an adult is diagnosed with diabetes  
Y = 0: event that an adult is diagnosed as not having diabetes  
P(Y=1|X=1) = 0.95 (correctly diagnosed as having diabetes)  
P(Y=1|X=0) = 0.02 (incorrectly diagnosed as having diabetes)  
*P(Y=1)* = P(X=1)P(Y=1|X=1) + P(X=0)P(Y=1|X=0)  
```{r}
"Probability that the community health service will diagnose an adult as having diabetes"
0.08*0.95 + 0.92*0.02
```
(b) an adult diagnosed by the health service as having diabetes actually has the
disease.  
Using the same notations as in a)  
*P(Y=1 and X=1)* = P(X=1)P(Y=1|X=1)  
```{r}
0.08*0.95
```

5. A bag contains 5 coins, one of which has a head on both sides, another one has a tail
on both sides while the other 3 coins are normal (with equal probability of tail and
head). A coin is chosen at random from the bag and tossed 2 times. The number of
heads obtained is a random variable, say X.  
(a) What are the possible values of X? Also tabulate the pmf of X.
(Hint: The coin chosen is either normal (a) or with head (b) or tail (c) on both
sides. Find the conditional probability of X in each of these three cases (a), (b)
and (c). Then use the law of total probability and multiplication rule to find the
pmf of X.)  
Suppose:  
Coin 1 is all-heads, probability of Heads p1 = 1  
Coin 2 is all-tails, probability of Heads p2 = 0  
Coin 3, 4, 5 are normal, probabilities of Heads p3 = p4 = p5 = 0.5  
Ai: event that coin i is selected (P(Ai) = 1/5)  
*Possible values of X* = 0, 1, 2  
*PMF of X:*  
P(X=0) = P(A1)P(X=0|A1) + P(A2)P(X=0|A2) + ... + P(A5)P(X=0|A5) = P(A1) {P(X=0|A1) + P(X=0|A2) + ... + P(X=0|A5)}  
** P(A1) = 1/5  
** Given coin 1 selected, X ~ Bi(n=2, p=p1) -> calculate P(X=0|A1)  
** Given coin 2 selected, X ~ Bi(n=2, p=p2) -> calculate P(X=0|A2)  
** Given coin 3 or 4 or 5 selected, X ~ Bi(n=2, p=p3) -> calculate P(X=0|A3) = P(X=0|A4) = P(X=0|A5)  
```{r}
p1 <- 1
p2 <- 0
p3 <- 0.5
px01 <- dbinom(0, size=2, prob=p1) # P(X=0|A1)
px02 <- dbinom(0, size=2, prob=p2) # P(X=0|A2)
px03 <- dbinom(0, size=2, prob=p3) # P(X=0|A3)

"P(X=0)"
1/5 * (px01 + px02 + 3*px03)
```
Similarly,  
P(X=1) = P(A1)P(X=1|A1) + P(A2)P(X=1|A2) + ... + P(A5)P(X=1|A5) = P(A1) {P(X=1|A1) + P(X=1|A2) + ... + P(X=1|A5)}  
** P(A1) = 1/5  
** Given coin 1 selected, X ~ Bi(n=2, p=p1) -> calculate P(X=1|A1)  
** Given coin 2 selected, X ~ Bi(n=2, p=p2) -> calculate P(X=1|A2)  
** Given coin 3 or 4 or 5 selected, X ~ Bi(n=2, p=p3) -> calculate P(X=1|A3) = P(X=1|A4) = P(X=1|A5)  
```{r}
px11 <- dbinom(1, size=2, prob=p1) # P(X=1|A1)
px12 <- dbinom(1, size=2, prob=p2) # P(X=1|A2)
px13 <- dbinom(1, size=2, prob=p3) # P(X=1|A3)

"P(X=1)"
1/5 * (px11 + px12 + 3*px13)
```
Similarly again,  
P(X=2) = P(A1)P(X=2|A1) + P(A2)P(X=2|A2) + ... + P(A5)P(X=2|A5) = P(A1) {P(X=2|A1) + P(X=2|A2) + ... + P(X=2|A5)}  
```{r}
pa1 = 1/5 # P(A1)
px21 <- dbinom(2, size=2, prob=p1) # P(X=2|A1)
px22 <- dbinom(2, size=2, prob=p2) # P(X=2|A2)
px23 <- dbinom(2, size=2, prob=p3) # P(X=2|A3)

"P(X=2)"
pa1 * (px21 + px22 + 3*px23)
```
Summary,  
$$
p_X = \left\{ \begin{array}{cl}
0.35 & \ ; \ X = 0 \\
0.3 & \ ; \ X = 1 \\
0.35 & \ ; \ X = 2 \end{array} \right.
$$

(b) Calculate E(X) and Var(X).  
```{r}
"E(X)"
x_mean <- 0.35*0 + 0.3*1 + 0.35*2
x_mean

"E(X^2)"
x_sq_mean <- 0.35*0^2 + 0.3*1^2 + 0.35*2^2
x_sq_mean

"Var(X)"
x_var <- x_sq_mean - x_mean^2
x_var
```

6. Among the 16 applicants for a job, 10 have university degrees. A sample of 3 applicants
are to be randomly chosen for interviews. Let X be the number of applicants in the
sample who have university degrees.  
(a) Give the name to the distribution of X if it has a name. Also specify the values
of all parameters involved in this distribution.  
**Hypergeometric distribution** with parameters:  
Population size (N) = 16  
Number of successes in population	(K) = 10  
Sample size (n) = 3  

(b) Find the probability that exactly 1 applicant in the sample has a university degree.  
**P(X=1)** = $\frac{{10 \choose 1}{6 \choose 2}}{{16 \choose 3}}$  
```{r}
choose(10, 1) * choose(6, 2) / choose(16, 3)
```

(c) Find the probability that at most 1 applicant in the sample has a university
degree.  
**P(X<=1)** = $P(X=0) + P(X=1)$  
```{r}
dhyper(0, m=10, n=6, k=3) + dhyper(1, m=10, n=6, k=3)
```

7. A moment-generating function of X is given by M(t) = e
t/(2 − e
t
).  
(a) Find the values of the mean, µ, and variance, σ
2
, for X.  
$$M_X(t) = E(e^{tX}) = 1 + tE(X) + \frac{t^2}{2!}E(X^2)+...+
\frac{t^k}{k!}E(X^k)+... \text{(Taylor series)}$$
$$E(X^k) = M_X^{(k)}(0) = \frac{d^k}{dt^k} M_X(t) |_{t=0}$$
**Mean**  
$$\mu = E(X) = \frac{d}{dt} M_X(t) |_{t=0} = \frac{2e^t}{(2-e^t)^2} \bigg|_{t=0} = 2$$
**Variance**  
$$E(X^2) = \frac{d^2}{dt^2} M_X(t) |_{t=0} = 2 \frac{2e^t + e^{2t}}{(2-e^t)^3} \bigg|_{t=0} = 6$$
$$\sigma^2 = E(X^2) - [E(X)]^2 = 6 - 2^2 = 2$$
(b) Calculate Pr{X ≥ 4}.  
Consider MGF of a Geometric distribution, with parameter p:  
$$
M_{Geom}(t) = \frac{pe^t}{1 - (1-p)e^t}
$$
Let $p = \frac{1}{2}$:  
$$
M_{Geom}(t) = \frac{\frac{1}{2}e^t}{1 - (1-\frac{1}{2})e^t} = \frac{e^t}{2 - e^t} = M_X(t)
$$
This means X ~ Geom(p = 1/2)  
**Pr{X ≥ 4}** = $1 - Pr(X < 4) = (1-p)^3$  
```{r}
(1-1/2)^3
```

