---
title: "MAST90105: Lab and Workshop 3"
output: html_notebook
---


# Lab  

1. An urn contains 17 balls marked LOSE and 3 balls marked WIN. You and an opponent
take turns selecting at random a single ball from the urn without replacement. The
person who selects the third WIN ball wins the game. It does not matter who selected
the first two WIN balls.  
a. If you draw first, find the probability that you win the game on your second draw.  
b. If you draw first, find the probability that your opponent wins the game on his
second draw.  
c. If you draw first, the probability that you win can be found from
P(You win if you draw first) = X
9
k=1

3
2

17
2k−2


20
2k
 ×
1
20 − 2k
(Why?)
Note: You could win on your second, third, fourth, ..., or tenth draw, not on your
first.  
**Hypergeometric distribution**, with parameters N = 20, K = 3, n = 2k  

d. If you draw second, the probability that you win can be found from
P(You win if you draw second) = X
9
k=1

3
2

17
2k−1


20
2k+1 ×
1
19 − 2k
. (Why?)  



e. Based on your results in (c) and (d), would you prefer to draw first or second?
Why?  

2. Suppose in a lot of 100 fuses there are 20 defective ones. A sample of 5 fuses are
randomly selected from the lot without replacement. Let X be the number of defective
fuses found in the sample.  
a. Find the probability P(X = 0),  
b. The cumulative probability P(X ≤ 3),  
c. The mean or expectation of X, E(X),  
d. The second moment of X, E(X2
),  
e. The variance of X, V ar(X),  
f. A probability bargraph for the pmf  
```{r}
x <- 0:5
y <- dhyper(x, m=20, n=20, k=5)
plot(x, y)
```

3. An urn contains n balls numbered from 1 to n. A random sample of n balls is selected
from the urn, one at a time. A match occurs if ball numbered i is selected on the ith
dr
• If the draws are done with replacement, it can be shown that
P(at least one match) = 1 −

1 −
1
n
n
.
• If the draws are done without replacement, it can be shown that
P(at least one match) = 1 −

1 −
1
1! +
1
2! −
1
3! + · · · +
(−1)n
n!

= 1−
Xn
k=0
(−1)k
k!
.  
a. For each value of n given in the following table, find P(at least one match) and
write down the results at appropriate places of the table.  
**Converge to e^k**
```{r}
match.f <- function(n, simsize, rep = TRUE) {
freq = 0
for (i in 1:simsize) {
sam = sample(1:n, size = n, replace = rep)
freq = freq + (sum(sam == 1:n) >= 1)
}
freq/simsize
}

match.f(n = 1, simsize = 1000, rep = TRUE)
```

b. We can also use R to simulate the processes of drawing n balls with or without
replacement from a set of n balls numbered from 1 to n. We can then simulate
the probability of at least one match using the relevant relative frequencies.
• Create a function match.f in R-Studio by writing a script and executing it
as follows.











